\section{Plan}
The following section summarizes and introduces the approaches we chose for the scope of this project. Each subsection describes the plan of one student of the team.

\subsection{Dataset and Preprocessing}
For all of the following approaches we plan to use the UNSW-NB15 \cite{UNSW} dataset, since the papers we evaluated for our project introduced UNSW-NB15.

\nextparagraph
To prepare the training set, as described in Section~\ref{sec:dataset}, we aim to normalize the input data with the help of min-max scaling to make the heterogenous features of the dataset more comparable so that no feature is favoured compared to flags, for instance. After that we will employ Synthetic Minority Oversampling Technique (SMOTE) to address the imbalance problem of favouring the majority class and enable our models to handle minority classes appropriately. With this baseline, the data can be used to train different models or use another preprocessing method to prepare for optimal attack classification.

\subsection{K-means-based Approach - Lisa Tschöscher}
In this project I will focus on applying k-means clustering on the data set, to recognize attacks in network traffic. The approach will be similar like in \cite{younisse2023empirical}.
In order to apply  k-means clustering on the chosen dataset the data is going to be preprocessed. Therefore categorical features like IP addresses, protocol, etc. are encoded and normalized (via min-max scaling) because k-means clustering needs numerical values. The training of the model will be mainly done with offline clustering due to the fact that the chosen dataset only provides static data. In addition I will try to simulate streaming data via streaming the dataset one data point at a time into the algorithm. This is done in order to test the online k-means clustering approach of the paper.
With clustering the input data is going to be partitioned into different clusters which indicate whether a data row is normal traffic or attack traffic. Using performance measures like F1-score and precision the results will be evaluated.

\subsection{SVM-based Approach - Laura Hofer}
Inspired from Turukmane and Devendiran \cite{SVM}, I'v decided to keep some of the basic architecture of the M-MultiSVM, but replace the used methods with less mathematical ones.

\nextparagraph
Since our group decided to use the UNSW-NB15 \cite{UNSW} dataset, this project will only use one dataset compared to \cite{SVM} and focus more on a different data preprocessing  approach and another hybrid model.

\nextparagraph
After the complexity reduction with the SMOTE algorithm, I plan to apply feature selection with the help of feature importance based on Random Forest to reduce noise and prepare the data for SVM. The inspiration for this combination of SVM and Random Forest stems from Dey et al. \cite{dey2023new}. Additionally, for tuning the SVM hyperparameter I will use the values described in \cite{dey2023new} to keep the optimization more simple than the Mud Ring optimization.

\subsection{Random Forest-based Approach - Thomas Wobak}
For my part of the project, I will implement a Random Forest (RF) classifier to detect network intrusions, drawing direct inspiration from the methodology proposed by \cite{intrusion_detection_combined_random_forest}. In their research, they successfully combined an enhanced Random Forest algorithm with the SMOTE oversampling technique to address the severe data imbalance often found in network security datasets, achieving high classification accuracy for both majority and minority classes.

\nextparagraph
Using our shared preprocessing pipeline, the input to the RF model will consist of the normalized and SMOTE-balanced UNSW-NB15 feature set. While \cite{wu2022intrusion} evaluated their model on the NSL-KDD dataset, I will adapt their approach to our chosen UNSW-NB15 data. During the training phase, I will utilize the ensemble nature of the Random Forest to handle the high dimensionality of the network traffic parameters. Furthermore, I plan to explore their proposed enhancements, such as evaluating individual decision trees based on the Area Under the Curve (AUC) index and filtering them based on structural similarity to optimize the forest's overall predictive voting process.

\nextparagraph
The model will output a classification prediction distinguishing between benign traffic and specific cyber attack signatures. Finally, the evaluation phase will measure the model's predictive capabilities using standard classification metrics, including accuracy, precision, recall, and the F1-score, directly mirroring the evaluation criteria used in the referenced study. By comparing these results against the clustering and SVM approaches developed by my peers, we will draw a comprehensive conclusion regarding the effectiveness of ensemble learning in this context.
