\section{Related Work}
Since numerous approaches to train ML learning models have been discussed in recent years, multiple surveys and literature reviews summarize different approaches and models and compare them \cite{abdulganiyu2023systematic} \cite{al2021intrusion} \cite{saranya2020performance} \cite{sowmya2023comprehensive}. Due to this variety of models and previous evaluations of different ML methods, this related work section focuses on concrete approaches that our group aims to evaluate and implement in parts.

\subsection{Mud Ring Assisted Multilayer Support Vector Machine (M-MultiSVM)}
Turukmane and Devendiran \cite{SVM} introduce a hybrid machine learning model called 'Mud Ring
assisted multilayer support vector machine (M-MultiSVM)', which addresses some limitations of ML models for IDS, which are also mentioned in \cite{al2021intrusion}. Namely, the authors apply Advanced Synthetic Minority Oversampling Technique (ASMOT) to address the class imbalance problem, favouring the majority class, and handle null values as well as unnormalized data with a Min-Max normalization beforehand. After that, the feature extraction is supported by Modified Singular value decomposition (M-SvD) to filter for basic, content and traffic features from the balanced input data. To further determine the optimal features, the Opposition learning-based Northern Goshawk optimization (ONgO) is applied, where necessary values for the attack classification are selected as well as improving the speed of convergence with Opposition-based learning. Finally, the hybrid Mud ring assisted multilayer support vector machine model can be trained with optimized, balanced and selected features. The classifier uses the traditional SVM techniques combined with an optimized hyperparameter created with the Mud Ring optimization. With a combination of the CIC-IDS 2018 and UNSW-NB15 dataset, the trained model is compared to standard ML models such as XGBoost or Logical Regression by evaluating performance measures such as accuracy, precision, recall, F1 score, False positive rate (FPR), Matthews correlation coefficient (MCC), Detection Rate, False Alarm Rate (FAR) and Specificity. The results show, that M-MultiSVM outperforms other approaches for almost every performance measure \cite{SVM}.


\subsection{An Empirical Study on Utilizing Online K-Means Clustering for Intrusion Detection Purposes}
This paper \cite{younisse2023empirical} addresses the problem of intrusion detection in network traffic. It focuses on two main aspects: \emph{online streaming environments} and \emph{highly imbalanced datasets}.
The authors emphasize the challenge, that traditional machine learning methods require labeled data and access to the full dataset, whereas in real network traffic, data often arrives sequentially and without labels.
The proposed approach is an unsupervised learning method using online K-Means clustering, which processes data sequentially.
Similar data points are grouped using \emph{Euclidean distance}.The model dynamically creates cluster centers and assigns new data points to the nearest center.

\nextparagraph
This approach enables real-time operation in environments where data arrives sequentially. For evaluation, the authors use the \emph{Kitsune dataset (IoT network traffic)} containing different attack types and 116 features. The data is first normalized via \emph{min-max normalization} (0–1 scaling) during preprocessing. Then the online K-Means algorithm is applied. The steps of the algorithm are:

\begin{enumerate}
    \item Initialization of the first $k$ incoming data points as initial cluster centers.
    \item Processing of each new data point:
    \begin{itemize}
        \item Calculate the Euclidean distance $D(v,C)$ between the new point $v$ and all existing cluster centers $C$.
        \item Determine the probability $p = \min(D(v,C)^2 / f_r, 1)$ of creating a new cluster center for $v$, where $f_r$ is a parameter controlling cluster growth.
        \item With probability $p$, $v$ becomes a new cluster center; otherwise, assign $v$ to the nearest existing center.
        \item Updating the cluster – the assigned cluster center may be updated incrementally if new points are added.
    \end{itemize}
    \item Output: clusters grow dynamically over time.
\end{enumerate}

\nextparagraph
The key findings of this study in include that, online K-Means performs almost as well as offline K-Means. Furthermore the authors recognized that normalization significantly improves the performance. Using different performance measure the following results were extracted. The cluster purity for both normal and attack traffic was measured at 99\%. The F1 score reaches 99\% for normal behavior and 94\% for attacks.

\nextparagraph
This approach can also be applied to the chosen \emph{UNSW-NB15} dataset, as both datasets represent network traffic, contain different behaviors, and have enough features for K-Means clustering \cite{younisse2023empirical}.